### 2020-02-22 doc/icfp20/counterpoint.tex
improve counterpoint section

The beginning of the music should express perfection.
As we saw in Section~\ref{sec:music}, there are three intervals that
are classified as perfect: the unison, the 5th, and the 8th.
The first interval of the music must then be one of these intervals.
In our formalization, we implement this rule as the
\texttt{checkBeginning} function, which reports an error
\texttt{not158 i} when the first interval \texttt{i} is an invalid one.
Since Agda does not have exceptions, we turn the error into an option
value by wrapping it around the \texttt{just} constructor.

@@

The middle bars of the music should maintain harmonical consonance
and independence of melodic lines.
As we saw before, consonant intervals include the unison, 3rd, 5th,
6th, and 8th, and among these, the unison is clearly an obstacle to
distinguish between the two lines of music.
Therefore, the middle bars must consist of the latter four intervals.
We encode this rule as the \texttt{checkIntervals} function, which
returns a list of errors correponding to the occurrences of dissonant
intervals and unisons.

@@

The independence of melodic lines is also affected by \emph{motion},
i.e., the way one interval moves to another interval.
As can be seen from Figure~\ref{fig:motion} (left), there are four
kinds of motion: \emph{contrary} (two lines go in different directions),
\emph{parallel} (two lines go in the same direction by the same
distance), \emph{similar} (two lines go in the same direction),
and \emph{oblique} (one line plays the same note).
Among these variants, parallel and similar motion makes it harder
to distinguish between two voices, especially when the second
interval is a perfect one, as in Figure~\ref{fig:motion} (right).
To avoid problematic motion, we define the \texttt{checkMotion}
function, which lists all invalid motion approaching a perfect interval.

@@

With this record type, we can show that the counterpoint composed
for the Frog's song is correct, since the music inhabits the
\texttt{FIrstSpecies} type.

@@

We next discuss a more complex variant of counterpoint, called the
second species.
In second species, we set \emph{two} half notes against every cantus
firmus note.
Figure~\ref{fig:ss} is an example of such two-against-one counterpoint,
again composed for the Frog's song.
As we can see from this example, the first and last bars may have
a different structure from middle bars.
More specifically, it is encouraged to begin the counterpoint line with
a half rest and end with a whole note.
Therefore, in our implementation, we reuse the \texttt{PitchInterval} 
type for the first and last bars, and define a new type
\texttt{PitchInterval2} for middle bars:

@@

The rhythmic structure of secound species counterpoint gives rise to
the distinction between \emph{strong beats} (the first interval in a bar)
and \emph{weak beats} (the second interval).
Roughly speaking, strong beats in middle bars are constrained by the
same rules as in first species.
That is, they must all be consonant, non-unison intervals, and in the 
case of perfect intervals, the motion from the preceding interval (on
the weak beat) must be either contrary or oblique.

Weak beats, on the other hand, are constrained in a looser manner.
First, they are allowed to be dissonant if they are created by a
\emph{passing tone}, i.e., a note in the middle of two step-wise
motions in the same direction (as in bars 4-5 of Figure~\ref{fig:ss}).
Second, they are allowed to be a unison if they are left by step in the
opposite direction from their approach (as in bars 5-6 of Figure~\ref{fig:ss}).

With these rules in mind, let us look at the definition of
\texttt{SecondSpecies}, a record type inhabited by correct second
species counterpoint:

@@

\noindent As in \texttt{FirstSpecies}, we have three fields holding
the musical content, followed by five fields carrying the proofs of
various properties.
Among the latter group of fields, \texttt{beginningOK} is defined
analogously to the corresponding field of \texttt{FirstSpecies};
the only difference is that the interval must not be a unison,
as it prevents the listener from recognizing the beginning of the
counterpoint line.
The next two fields, \texttt{strongBeatsOK} and \texttt{weakBeatsOK},
guarantee the correctness of strong and weak beats, respectively.
What follows is \texttt{motionOK}, which converts a list of
\texttt{PitchInterval2} into a list of \texttt{PitchInterval}, and checks
motion accross bars.
Lastly, we have \texttt{endingOK}, which forces the last two intervals
to have a cadence structure.

Using \texttt{SecondSpecies}, we can show that our second species
counterpoint for the Frog's song is correct.

>>>>

The beginning of the music should express perfection.
As we saw in Section~\ref{sec:music}, there are three intervals that
are classified as perfect: the 1st (more commonly called the
\emph{unison}), the 5th, and the 8th.
The first interval of the music must then be one of these intervals.
In our formalization, we implement this rule as the
\texttt{checkBeginning} function, which reports an error
\texttt{not158 i} when the first interval \texttt{i} is an invalid one.
Since Agda does not have exceptions, we turn the error into an option
value by wrapping it around the \texttt{just} constructor.

@@

The middle bars of the music should maintain harmonical consonance
and independence of melodic lines.
As we saw before, consonant intervals include the unison,
3rd\footnote{We use ``3rd'' to mean both the major and minor variants
of the 3rd interval, and similarly for the 6th.}, 5th, 6th, and 8th,
and among these, the unison is clearly an obstacle to distinguishing
between the two lines of music.
Therefore, the middle bars must consist of the latter four intervals.
We encode this rule as the \texttt{checkIntervals} function, which
returns a list of errors correponding to the occurrences of dissonant
intervals and unisons.

@@

The independence of melodic lines is also affected by \emph{motion},
i.e., the way one interval moves to another interval.
As can be seen from Figure~\ref{fig:motion} left, there are four
kinds of motion: \emph{contrary} (two lines go in different directions),
\emph{parallel} (two lines go in the same direction by the same
distance), \emph{similar} (two lines go in the same direction),
and \emph{oblique} (one line plays the same note).
There is a consensus that approaching a perfect interval by parallel
or similar motion (as in Figure~\ref{fig:motion} right) destroys the
independence of lines.
To rule out this, we define the \texttt{checkMotion} function, which
lists all occurrences of such invalid motion.

@@

With this record type, we can show that the counterpoint in
Figure~\ref{fig:fs} is correct, since the music is an inhabitant of
the \texttt{FIrstSpecies} type.

@@

We next discuss a more complex variant of counterpoint, called the
second species.
In second species, we set \emph{two} half notes against every cantus
firmus note.
Figure~\ref{fig:ss} is an example of such two-against-one counterpoint,
again composed for the Frog's song.

In second species, the first and last bars may have a different structure
from middle bars.
More specifically, it is encouraged to begin the counterpoint line with
a half rest and end with a whole note.
Therefore, in our implementation, we reuse the \texttt{PitchInterval} 
type for the first and last bars, and define a new type
\texttt{PitchInterval2} for middle bars:

@@

In second species counterpoint, some parts of the music are applied
the same rules as first species, while other parts are constrained by
new rules.

The beginning of the music is \emph{not} allowed to be the unison,
as it prevents the listener from recognizing the beginning of the
counterpoint line.
We restrict the first interval by defining a new error
\texttt{BeginningError2} and a function \texttt{checkBeginning2}
that works analogously to \texttt{checkBeginning} for first species.

\paragraph{Strong Beats}

The rhythmic structure of secound species counterpoint gives rise to
the distinction between \emph{strong beats} (the first interval in a bar)
and \emph{weak beats} (the second interval).
Strong beats in middle bars are constrained by the same rules as in
first species.
That is, they must all be consonant, non-unison intervals, and in the 
case of perfect intervals, the motion from the preceding interval (on
the weak beat) must be either contrary or oblique.
We encode these restrictions as the \texttt{checkStrongBeats}
function, which, as the \texttt{checkIntervals} function for first
species, returns a list of found errors.

\paragraph{Weak Beats}

Compared to strong beasts, weak beats are constrained in a looser
manner.
First, they are allowed to be dissonant if they are created by a
\emph{passing tone}, i.e., a note in the middle of two step-wise
motions in the same direction (as in bars 4-5 of Figure~\ref{fig:ss}).
Second, they are allowed to be the unison if they are left by step in
the opposite direction from their approach (as in bars 5-6 of
Figure~\ref{fig:ss}).
We implement these rules as the \texttt{checkWeakBeats} function,
which checks the validity of every three successive beats in the
middle bars.

Parallel and similar motion towards a perfect interval is prohibited
across bars.
We enforce this rule using the \texttt{checkMotion2} function,
which uses \texttt{expandPitchIntervals2} to convert a list of
\texttt{PitchInterval2} into a list of \texttt{PitchInterval}, and
checks every pair of weak beat and strong beat intervals.

Now we define \texttt{SecondSpecies}, a record type inhabited by
correct second species counterpoint.
As in \texttt{FirstSpecies}, we have three fields holding the musical
content, followed by five fields carrying the proofs of various
properties.

@@

Using \texttt{SecondSpecies}, we can show that the second species
counterpoint in Figure~\ref{fig:ss} is correct.
